---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r preprocessing}
activity <- read.csv('activity.csv')
```


## What is mean total number of steps taken per day?

Calculate total number of daily steps by adding the steps of all 5 minute intervals in a day. Plot a histogram of this total number of steps.
```{r steps}
dailySteps <- tapply(activity$steps, activity$date, sum, simplify=TRUE)
hist(dailySteps, main="Total number of steps per day", xlab="steps")
```

The mean and median number of steps are calculated, after removing the missing values.
```{r mean_and_median}
mean(dailySteps, na.rm = TRUE)
median(dailySteps, na.rm = TRUE)
```


## What is the average daily activity pattern?

Calculate mean of number of steps for each 5 minute interval, averaged over all days.
```{r daily_activity}
intervalSteps <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE, simplify=TRUE)
interval <- unique(activity$interval)
plot(interval, intervalSteps, type='l', main='Average steps per 5 minute interval', ylab="steps")
```

Find largest value and its index, then find the corresponding interval.
```{r daily_largest}
maxsteps <- max(intervalSteps, na.rm = TRUE)
maxinterval<- names(intervalSteps)[intervalSteps==maxsteps]
cat("Interval", maxinterval, "has the largest average number of steps:", maxsteps, "\n")
```

## Imputing missing values

Impute missing values with average number of steps of that interval over all days (as calculated above).

```{r imputing}
na_values <- length(activity$steps[is.na(activity$steps)])
cat("Total number of rows with missing values for steps is",na_values, "out of a total of",length(activity$steps), "measurements.")
```

Impute missing values with average number of steps of that interval over all days (as calculated above).

```{r imputing2}
activity2 <- activity # create new dataset, initially identical to old one
for(i in 1:nrow(activity)) { # loop over rows of original dataset
    row <- activity[i,]
    if( is.na(row[,1]) ){ # if number of steps is NA
      ival = row[,3] # interval value
      iival = which(interval==ival) # interval index in the vector of means per interval
      activity2[i,1] <- intervalSteps[iival] # actual imputing of values
    }
}
```

Re-calculate total number of steps per day and redo histogram using the imputed dataset.
```{r total_steps_after_imputing}
dailySteps2 <- tapply(activity2$steps, activity2$date, sum, simplify=TRUE)
hist(dailySteps2, main="Total number of steps per day after imputing", xlab="steps")
```

The mean and median number of steps after imputing
```{r mean_and_median_after_imputing}
mean(dailySteps2, na.rm = TRUE) # na.rm=TRUE actually not needed anymore
median(dailySteps2, na.rm = TRUE)
```
The mean and the median after imputing do not change much from the ones we calculated before imputing.

## Are there differences in activity patterns between weekdays and weekends?

First convert days to weekdays, then to two factors. Add those factors to the data set.

```{r weekdays_weekends}
what_day <- weekdays(as.Date(activity2$date))
what_day[what_day=="Sunday" | what_day=="Saturday"] = "Weekend"
what_day[what_day!="Weekend"] = "Weekday"
activity2$day <- what_day
#
intervalStepsWeekend <- tapply(activity2$steps[activity2$day=="Weekend"], activity2$interval[activity2$day=="Weekend"], mean, na.rm = TRUE, simplify=TRUE)
intervalStepsWeekday <- tapply(activity2$steps[activity2$day=="Weekday"], activity2$interval[activity2$day=="Weekday"], mean, na.rm = TRUE, simplify=TRUE)

par(mfrow=c(2,1))
plot(interval, intervalStepsWeekend, type='l', main='weekend', ylab="steps", ylim=c(0.,250.))
plot(interval, intervalStepsWeekday, type='l', main='weekday', ylab="steps", ylim=c(0.,250.))
```

